---
title: "Random Forests and Boosted Regression Trees - Homework"
output: 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
editor_options: 
  chunk_output_type: inline

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(include = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = 'center')

```

<br>

**This assignment is due on December 7.**

**Please use the corresponding R Markdown file to save your results in an HTML file. Then upload the HTML file to learn.wu.ac.at.**

<br>

# Input data

Install and load the required packages.

```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

# Load packages
library(plyr)
library(tidyverse)
library(data.table)
library(fst)
library(ranger)
library(xgboost)
library(caret)
library(viridis)

```

Adjust the data path and load the input data.

```{r}

# Set the path to the data
data_path = "/Users/annevalder/Desktop/UNI /WU WIEN/WS 20:21/Data Science/Exercises/RCTCV_A2"

# Read input data on municipal level for the whole country
col_input = read_fst(file.path(data_path, "data", "colombia_input.fst"))

# Read input data on grid level for the department Quindío
quindio_grid = read_fst(file.path(data_path, "data", "quindio_input_grid.fst"))

```

# Data Visualization and Exploratory Data Analysis

## Exercise 1

Use the input data for Colombia on the municipal level (colombia_input.fst) for this exercise

### 1.1

Create a plot that shows the distribution of the popualtion density (raster_pop_100). Briefly describe what you see.
```{r}

col_input %>% ggplot(aes(x=raster_pop_100)) + geom_density() +

```



### 1.2

Create a plot that visualizes the relationship between the population density (raster_pop_100) and the light intensity recorded at night (night_lights_100). Combine two suitable geom functions in one plot. Briefly describe what you see.

plot(col_input$raster_pop_100,col_input$night_lights_100)


## Exercise 2

Use the grid-level input data for Quindío (quindio_input_grid.fst) for this exercise.

### 2.1

Create a plot the shows the distribution of the light intensity at night (night_lights_100). Include information on whether or not a grid lies inside a protected area (protected_areas_100). Briefly describe what you see.

Hint: Transform the variable protected_areas_100 to character inside the ggplot or the geom function.

### 2.2

Create a plot that visualizes the relationship between the light intensity at night (night_lights_100) and the slope (hydro_slo_100). What could be a problem with a standard scatterplot? Try to solve this issue and briefly describe what you see.

# Random Forests and Boosted Regression Trees

## Exercise 3

### 3.1

Use the input data for Colombia on the municipal level (colombia_input.fst) to train a Random Forest. Use the log of the population density (raster_pop_100) as the dependent variable. Tune one of the parameters in the Random Forest. Briefly describe the parameter you are tuning and explain why you have decided for this parameter.

### 3.2

Use the best Random Forest model from the tuning exercise to make a prediction of the grid-level population density in Quindío. Make sure that the sums of the grid-level predictions match the official numbers on the municipal level.

## Exercise 4

### 4.1

Use the input data for Colombia on the municipal level (colombia_input.fst) to train a Boosted Regression Tree. Use the log of the population density (raster_pop_100) as the dependent variable. Tune one of the parameters in the Boosted Regression Tree. Briefly describe the parameter you are tuning and explain why you have decided for this parameter.

### 4.2

Use the best Boosted Regression Tree model from the tuning exercise to make a prediction of the grid-level population density in Quindío. Make sure that the sums of the grid-level predictions match the official numbers on the municipal level.

# Visualization of Results

## Exercise 5

Create a plot showing the following three maps of the population density in Quindío side by side: one based on the input data, one based on your Random Forest prediction, and one based on your Boosted Regression Tree prediction. Try to make sure that the maps give a good picture of the differences in population density, both between the different data sets as well as between the different parts of the department.





<br>
<br>
<br>
<br>

